Entropy-SGD: Biasing Gradient Descent Into Wide Valleys

### 2017-01-12 doc/chaudhari.choromanska.ea.iclr17.tex
pulling back updates from overleaf

We first provide a simple intuition for the concept of local entropy of an energy landscape. Consider a cartoon energy landscape in Fig.~\ref{fig:entropyfig} where the x-axis denotes the configuration space of the parameters. We have constructed two local minima: a shallower although wider one at $x_{\trm{robust}}$ and a very sharp global minimum at $x_{\trm{non-robust}}$. Under a Bayesian prior on the parameters, say a Gaussian of a fixed variance at locations $x_{\trm{robust}}$ and $x_{\trm{non-robust}}$ respectively, the wider local minimum has a higher marginalized likelihood than the sharp valley on the right.

@@

We can show that $\entropysgd$ results in a smoother loss function and obtains better generalization error than the original objective~\eqref{eq:dnn_objective}. With some overload of notation, we assume that the original loss $f(x)$ is $\b$-smooth, i.e., for all $x,y \in \reals^n$, we have $\norm{\nabla f(x) - \nabla f(y)} \leq \b\ \norm{x-y}$. We additionally assume for the purpose of analysis that no eigenvalue of the Hessian $\nabla^2 f(x)$ lies in the set $[-2 \g-c, c]$ for some small $c > 0$.
%
\begin{lemma}
\label{lem:smoothness_reduction}
The objective $F(x, \g;\ \Xi)$ in~\eqref{eq:entropysgd_objective} is $\f{\a}{1 + \g^{-1}\ c}$-Lipschitz and $\f{\b}{1 + \g^{-1}\ c}$-smooth.
\end{lemma}
Please see Sec.~\ref{s:app:proofs} for the proof. The local entropy objective is thus smoother than the original objective.
%
Let us now obtain a bound on the improvement in generalization error. We denote an optimization algorithm, viz., SGD or $\entropysgd$ by $A(\Xi)$, it is a function of the dataset $\Xi$ and outputs the parameters $x^*$ upon termination. Stability of the algorithm~\citep{bousquet2002stability} is then a notion of how much its output differs in loss upon being presented with two datasets $\Xi$ and $\Xi'$ that differ in at most one sample,
$$
    \sup_{\xi\ \in\ \Xi\ \cup\ \Xi'}\ \sqbrac{ f \rbrac{A(\Xi), \xi} - f \rbrac{A(\Xi'), \xi}} \leq \e.
$$
\citet{hardt2015train} connect uniform stability to generalization error and show that an $\e$-stable algorithm $A(\Xi)$ has generalization error bounded by $\e$, i.e., if $A(\Xi)$ terminates with parameters $x^*$,
$$
    \abs{ \E_{\Xi} \rbrac{R_\Xi(x^*) - R(x^*)} } \leq \e;
$$
where the left hand side is the generalization error: it is the difference between the empirical loss $R_\Xi(x) := N^{-1} \sum_{k=1}^N\ f(x, \xi_k)$ and the population loss $R(x) := \E_{\xi}\ f(x, \xi)$.
%
We now employ the following theorem that bounds the stability of an optimization algorithm through the smoothness of its loss function and the number of iterations on the training set.
\begin{theorem}[\citet{hardt2015train}]
\label{thm:generalization_bound_nonconvex}
For an $\a$-Lipschitz and $\b$-smooth loss function, if SGD converges in $T$ iterations on $N$ samples with decreasing learning rate $\eta_t \leq 1/t$ the stability is bounded by
\begin{align*}
    \e
    %&\leq \f{1}{N-1}\ \rbrac{1 + \f{1}{\b c}}\ \rbrac{2 c \a^2}^{\f{1}{1 + \b c}}\ T^{\f{\b c}{1 + \b c}}\\
    &\lessapprox \f{1}{N}\ \a^{1/(1 + \b)}\ T^{1 - 1/(1 + \b)}.
\end{align*}
\end{theorem}
%
Using Lem.~\ref{lem:smoothness_reduction} and Thm.~\ref{thm:generalization_bound_nonconvex} we have
\begin{equation}
    \e_{\ \entropysgd} \lessapprox \rbrac{\a\ T^{-1}}^{\rbrac{1 - \f{1}{1 + \g^{-1}c}}\ \b}\ \e_{\ \textrm{SGD}},
    \label{eq:stability_bound}
\end{equation}
which shows that $\entropysgd$ generalizes better than SGD for all $T > \a$ if they both converge within $T$ parameter updates. \pc{However, from our experiments, $\entropysgd$ converges in roughly $L$ times fewer parameter updates than SGD, using this we can draw out the dependency on the number of Langevin iterations to get
$$
    \e_{\ \entropysgd} \lessapprox L^{-\f{c\ \b}{\g}}\ \e_{\ \textrm{SGD}}.
$$
As an aside, it is easy to see from the proof of Lem.~\ref{lem:smoothness_reduction} that for a convex loss function $f(x)$, adding the local entropy term does not change the minimizer of the original problem.

\begin{remark}
The above analysis hinges upon an assumption that the Hessian $\nabla^2 f(x)$ does not have eigenvalues in the set $[-2\g-c, c]$ for a constant $c > 0$. This is admittedly unrealistic, for instance, the eigenspectrum of the Hessian at a local minimum in Fig.~\ref{fig:lenet_hessian} has a large fraction of its eigenvalues near zero. Let us however remark that the bound on uniform stability in Thm.~\ref{thm:generalization_bound_nonconvex} assumes global conditions on the smoothness of the loss function~\citep{hardt2015train}; one imagines that Eqn.~\ref{eq:stability_bound} remains qualitatively the same (with respect to $T$ in particular) even if this assumption is violated to an extent before convergence happens.
\end{remark}
}{}

@@

\textbf{Remark on the computational complexity:} Since $\entropysgd$ runs $L$ steps of SGLD before each parameter update, the effective number of passes over the dataset is $L$ times that of SGD or Adam for the same number of parameter updates. We therefore plot the error curves of $\entropysgd$ in Fig.~\ref{fig:mnist_test} and~\ref{fig:allcnn} against the ``effective number of epochs'', i.e. by multiplying the abscissae by a factor of $L$. (define $L = 1$ for SGD or Adam). This is thus a direct measure of wall-clock time, agnostic to the underlying hardware and implementation.

@@

Tuning the momentum in $\entropysgd$ was crucial to getting good results on RNNs. While the SGD baseline on $\ptblstm$ does not use momentum (and in fact, does not train well with momentum) we used a momentum of $0.5$ for $\entropysgd$. On the other hand, the baseline for $\charlstm$ was trained with Adam with $\b_1 = 0.9$ ($\b_1$ in Adam controls the moving average of the gradient) while we set $\b_1 = 0.5$ for $\entropyadam$. In contrast to this observation about RNNs, all our experiments on CNNs used a momentum of $0.9$. \pc{In order to investigate this difficulty, we monitored the angle between the local entropy gradient and the vanilla SGD gradient during training. This angle changes much more rapidly for RNNs than for CNNs which suggests a more rugged energy landscape for the former; local entropy gradient is highly uncorrelated with the SGD gradient in this case.}{}

@@

\todo{
\section{Connection to variational inference}
\label{s:app:connection_variational}

@@

Training Bayesian deep networks with SGLD, or other more sophisticated algorithms such as SGHMC~\citep{chen2014stochastic,neal2011mcmc} is difficult and the lack of such results in the literature corroborates our experimental results. Roughly speaking, local entropy is so effective because it operates on a transformation of the energy landscape that exploits local entropic effects. Conventional MCMC techniques such as SGLD or Nose'-Hoover thermostats~\citep{ding2014bayesian} can only trade energy for entropy via the temperature parameter which does not allow direct use of the geometric information of the energy landscape and does not help with narrow minima.
}

>>>>

We first provide a simple intuition for the concept of local entropy of an energy landscape. \pc{The discussion in this section builds upon the results of~\citet{baldassi2016unreasonable} and extends it for the case of continuous variables.}{} Consider a cartoon energy landscape in Fig.~\ref{fig:entropyfig} where the x-axis denotes the configuration space of the parameters. We have constructed two local minima: a shallower although wider one at $x_{\trm{robust}}$ and a very sharp global minimum at $x_{\trm{non-robust}}$. Under a Bayesian prior on the parameters, say a Gaussian of a fixed variance at locations $x_{\trm{robust}}$ and $x_{\trm{non-robust}}$ respectively, the wider local minimum has a higher marginalized likelihood than the sharp valley on the right.

@@

We can show that $\entropysgd$ results in a smoother loss function and obtains better generalization error than the original objective~\eqref{eq:dnn_objective}. With some overload of notation, we assume that the original loss $f(x)$ is $\b$-smooth, i.e., for all $x,y \in \reals^n$, we have $\norm{\nabla f(x) - \nabla f(y)} \leq \b\ \norm{x-y}$. We additionally assume for the purpose of analysis that no eigenvalue of the Hessian $\nabla^2 f(x)$ lies in the set $[-2 \g-c, c]$ for some small $c > 0$.
%
\begin{lemma}
\label{lem:smoothness_reduction}
The objective $F(x, \g;\ \Xi)$ in~\eqref{eq:entropysgd_objective} is $\f{\a}{1 + \g^{-1}\ c}$-Lipschitz and $\f{\b}{1 + \g^{-1}\ c}$-smooth.
\end{lemma}
Please see Sec.~\ref{s:app:proofs} for the proof. The local entropy objective is thus smoother than the original objective.
%
Let us now obtain a bound on the improvement in generalization error. We denote an optimization algorithm, viz., SGD or $\entropysgd$ by $A(\Xi)$, it is a function of the dataset $\Xi$ and outputs the parameters $x^*$ upon termination. Stability of the algorithm~\citep{bousquet2002stability} is then a notion of how much its output differs in loss upon being presented with two datasets $\Xi$ and $\Xi'$ that differ in at most one sample,
$$
    \sup_{\xi\ \in\ \Xi\ \cup\ \Xi'}\ \sqbrac{ f \rbrac{A(\Xi), \xi} - f \rbrac{A(\Xi'), \xi}} \leq \e.
$$
\citet{hardt2015train} connect uniform stability to generalization error and show that an $\e$-stable algorithm $A(\Xi)$ has generalization error bounded by $\e$, i.e., if $A(\Xi)$ terminates with parameters $x^*$,
$$
    \abs{ \E_{\Xi} \rbrac{R_\Xi(x^*) - R(x^*)} } \leq \e;
$$
where the left hand side is the generalization error: it is the difference between the empirical loss $R_\Xi(x) := N^{-1} \sum_{k=1}^N\ f(x, \xi_k)$ and the population loss $R(x) := \E_{\xi}\ f(x, \xi)$.
%
We now employ the following theorem that bounds the stability of an optimization algorithm through the smoothness of its loss function and the number of iterations on the training set.
\begin{theorem}[\citet{hardt2015train}]
\label{thm:generalization_bound_nonconvex}
For an $\a$-Lipschitz and $\b$-smooth loss function, if SGD converges in $T$ iterations on $N$ samples with decreasing learning rate $\eta_t \leq 1/t$ the stability is bounded by
\begin{align*}
    \e
    %&\leq \f{1}{N-1}\ \rbrac{1 + \f{1}{\b c}}\ \rbrac{2 c \a^2}^{\f{1}{1 + \b c}}\ T^{\f{\b c}{1 + \b c}}\\
    &\lessapprox \f{1}{N}\ \a^{1/(1 + \b)}\ T^{1 - 1/(1 + \b)}.
\end{align*}
\end{theorem}
%
Using Lem.~\ref{lem:smoothness_reduction} and Thm.~\ref{thm:generalization_bound_nonconvex} we have
\begin{equation}
    \e_{\ \entropysgd} \lessapprox \rbrac{\a\ T^{-1}}^{\rbrac{1 - \f{1}{1 + \g^{-1}c}}\ \b}\ \e_{\ \textrm{SGD}},
    \label{eq:stability_bound}
\end{equation}
which shows that $\entropysgd$ generalizes better than SGD for all $T > \a$ if they both converge in $T$ parameter updates. However, from our experiments, $\entropysgd$ converges in roughly $L$ times fewer parameter updates than SGD, using this we can draw out the dependency on the number of Langevin iterations to get
$$
    \e_{\ \entropysgd} \lessapprox L^{-\f{c\ \b}{\g}}\ \e_{\ \textrm{SGD}}.
$$
As an aside, it is easy to see from the proof of Lem.~\ref{lem:smoothness_reduction} that for a convex loss function $f(x)$, adding the local entropy term does not change the minimizer of the original problem.

\pc{
\begin{remark}
The above analysis hinges upon an assumption that the Hessian $\nabla^2 f(x)$ does not have eigenvalues in the set $[-2\g-c, c]$ for a constant $c > 0$. This is admittedly unrealistic, for instance, the eigenspectrum of the Hessian at a local minimum in Fig.~\ref{fig:lenet_hessian} has a large fraction of its eigenvalues near zero. Let us however remark that the bound on uniform stability in Thm.~\ref{thm:generalization_bound_nonconvex} by~\citet{hardt2015train} assumes global conditions on the smoothness of the loss function; one imagines that Eqn.~\ref{eq:stability_bound} remains qualitatively the same (with respect to $T$ in particular) even if this assumption is violated to an extent before convergence happens. Obtaining a rigorous generalization bound without this assumption would require a dynamical analysis of SGD and seems out of reach of the current literature.
\end{remark}
}{}

@@

\textbf{Remark on the computational complexity:} Since $\entropysgd$ runs $L$ steps of SGLD before each parameter update, the effective number of passes over the dataset is $L$ times that of SGD or Adam for the same number of parameter updates. We therefore plot the error curves of $\entropysgd$ in Figs.~\ref{fig:mnist_test},~\ref{fig:allcnn}, and~\ref{fig:rnn} against the ``effective number of epochs'', i.e. by multiplying the abscissae by a factor of $L$. (define $L = 1$ for SGD or Adam). This is thus a direct measure of wall-clock time, agnostic to the underlying hardware and implementation.

@@

Tuning the momentum in $\entropysgd$ was crucial to getting good results on RNNs. While the SGD baseline on $\ptblstm$ does not use momentum (and in fact, does not train well with momentum) we used a momentum of $0.5$ for $\entropysgd$. On the other hand, the baseline for $\charlstm$ was trained with Adam with $\b_1 = 0.9$ ($\b_1$ in Adam controls the moving average of the gradient) while we set $\b_1 = 0.5$ for $\entropyadam$. In contrast to this observation about RNNs, all our experiments on CNNs used a momentum of $0.9$. In order to investigate this difficulty, we monitored the angle between the local entropy gradient and the vanilla SGD gradient during training. This angle changes much more rapidly for RNNs than for CNNs which suggests a more rugged energy landscape for the former; local entropy gradient is highly uncorrelated with the SGD gradient in this case.

@@

\section{Connection to variational inference}
\label{s:app:connection_variational}

@@

Training Bayesian deep networks with SGLD, or other more sophisticated algorithms such as SGHMC~\citep{chen2014stochastic,neal2011mcmc} is difficult and the lack of such results in the literature corroborates our experimental results. Roughly speaking, local entropy is so effective because it operates on a transformation of the energy landscape that exploits local entropic effects. Conventional MCMC techniques such as SGLD or Nose'-Hoover thermostats~\citep{ding2014bayesian} can only trade energy for entropy via the temperature parameter which does not allow direct use of the geometric information of the energy landscape and does not help with narrow minima.
